Adaptive Waveform Inversion for Transmitted Wavefields

### 2024-08-30 repro/papers/awi/paper.tex
edits

While widely used in recent years, FWI exhibits a pathology when wave
velocities or their parametric equivalents are amongst the parameters
to be estimated: the excessive sensitivity of predicted waveforms to
changes in velocity leads to oscillations in the difference measures,
especially for the most commonly used mean-square measure. Local
optimization algorithms (relatives of Newton;'s method) are the only practical
methods to generate sequences of approximate optimal
parameters. However the oscillations in the difference measures lead
these sequences to become entrapped at near parameter distributions
far from the optimum, and thus do not extract the maximum information
from the data \cite[]{VirieuxOperto:09}.

@@

While these and other works discuss heuristic justifications for AWI and
many details of implementation, a clear explanation for its effectiveness is 
so far missing from the literature. In this short note, I show that
for active sensing of acoustic materials with unique arrival times of
transmitted waves
(that is, unique rays of geometric asymptotics between sources and
receivers), the AWI objective function is an asymptotic (in time
frequency) approximation to the
mean-square time-of-travel misfit. Thus AWI and travel time fitting
lead to similar material parameter estimates. Since the latter is
widely believed to be a near-optimal method for wave velocity
estimation, this demonstration explains roughly ``why AWI works'', at
least within this restricted scope.

@@

The version of AWI introduced by \cite{Warner:16} is based on an
active source acoustic model of data generation. Each acoustic source
is modeled an isotropic point radiator with known location $\bx_s$ and
wave form (``wavelet'') $w(t;\bx_s)$. Similarly, receivers are modeled
as isotropic point sensors, recording pressure at points ${\bf x}_r$. The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
For sake of brevity, define the model vector $m=(\kappa,\rho)$. The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which source and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.

Acoustic inversion, for present purposes, means: given source and receiver locations $(\bx_s,\bx_r)$, source wavelet
$w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
so that $F[m]w \approx d$. The
simplest version of FWI concretizes this task by asking for a model
$m$ minimizing the mean square error
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s} \int\,dt\, |(F[m]w)({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
\end{equation}
It is convenient to introduce the $L^2$ norm $\| \cdot \|$ as
shorthand for the right-hand side of the definition \ref{eqn:fwi}:
\[
  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]w-d\|^2.
\]
In the sequel I will use the $L^2$ norm notation to stand for
mean-square over whatever variables make sense.

Note that in actual data acquisition, recording occurs over finite
time intervals, so the integrals appearing in the definition
\ref{eqn:fwi} have finite upper and lower limits, as opposed to the
$\pm \infty$ limits implicit in the notation. It will be assumed that
wavefield energy is causal (zero before an onset time) and decays
rapidly enough as $t \rightarrow \infty$ that the difference between
finite and infinite range of integration can be ignored throughout
this discussion.

The wavelet $w$ is assumed known in this formulation of the
FWI problem. In fact, even if actual acoustic sources are
well-approximated by isotropic point radiators, their wave forms are
usually poorly determined and may reasonably be
included as  parameters to be estimated by inversion of the
data. However for the present discussion we will regard $w$ as known.

Practical versions of FWI include various types of regularization and
imposition of constraints on $m$, see the sources cited in the
Introduction. As mentioned there, local iterative optimization methods
are the only practical approach to minimizing $J_{\rm FWI}$ due to the
computational cost of approximating the forward map $F$ and related
operators, but these tend to stagnate at highly sub-optimal
(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
quite high quality, often available only at considerable cost in data
collection and processing.
%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m^*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa^*, \beta^*, w^*$: that is, $d = F[m^*]w^*$.

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources
$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
field. That is, the extended source depends on the receiver location
as well as the source location - so there is one acoustic system for
each source {\em and} receiver position - a lot of wave equations! If all
of the wavelets for each source are the same, that is, ${\bar w}({\bf
  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.

AWI assumes that the extended sources are time convolutions of the
(known) source $w$ with an {\em adaptive filter} $u({\bf
  x}_r,t;{\bf x}_s)$: that is,
\[
  \bar{w}({\bf x}_r,t;{\bf x}_s) = (u * w)({\bf  x}_r,t;{\bf x}_s)
  = (u({\bf  x}_r,\cdot;{\bf x}_s)*w)(t).
\]
The asterisk denotes convolution in time. That is,
$u*w$ denotes the result of convolving each trace in $u$ by $w$. I
will also use the asterisk to denote the tracae-by-trace convolution
of two data sets with the same set of source and receiver coordinates,
that is, the data set consisting of time convolutions of traces with the same source
and receiver coordinates. The meaning will be clear from context.

Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
\[
 \bar{F}[m]\bar{w} = \bar{F}[m](u* w) = (F[m]w)*u
\]
Since the known source $w$ is not adjusted during the inversion, 
introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w$:
\begin{equation}
  \label{eqn:sdef}
  (F[m]w)*u \equiv S[m]u
\end{equation}
\cite{Warner:16} assume (implicitly) that there is always an exact
solution of the extended inversion problem. That is, they assume that
for any $m$, there is an extended source $\bar{w}$, or equivalently an
adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
$S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
around, as will be seen below. Also, \cite{Warner:16} also formulate
AWI in terms of the inverse filter, that is, $u$ for which
$u*d = F[m]w$, and in fact recommend that definition. The present
discussion uses only the forward filter.

If $\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
${\bf x}_s,{\bf x}_r$. Such a kernel $u[m;d]$ is in the null space of
multiplication by $t$.

This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
\begin{equation}
  \label{eqn:tJz}
  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[m;d](\bx_r,t;\bx_s)|^2}{\int dt |u[m;d](\bx_r,t;\bx_s)|^2}\right)
\end{equation}
in which $u[m;d]$ is as defined above, that is, the solution of $S[m]u[m;d]
= d$.

Adaptive waveform inversion consists in iterative minimization of
$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta_0)$ of the
model, then update $m=(\kappa,\beta)$ repeatedly using a gradient
descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
and its gradient in detail, and their use in a quasi-Newton
optimization algorithm.

@@

It may not be obvious at first glance why this approach to estimating
$m$ should be any more effective than FWI. To see why AWI works, in at
least some settings, we assume that the spatial dimension is three,
and make use of the geometric asymptotics (ray theory) approximation
to the solution of the wave equation. This approximation is {\em
  local}: it applies when sources and receivers are sufficiently
close. We will discuss its global failure and the consequences for AWI
in the final section.

Transient acoustic sources have no DC (zero frequency) component and
are relatively small at low frequencies. One way to enforce these
characteristics is to assume that the wavelet $w$ is the time
derivative of another wavelet $v$, also transient ($v(t;{\bf x}_s)=0$
for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy
for wavelength: it is the reciprocal of RMS frequency of $v$.

The geometric asymptotics approximation for solution of the point
radiator problem \ref{eqn:awe} expresses the pressure field in terms
of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
travel time $\tau[m]({\bf x}_s,{\bf x}_r)$, and implies that
\begin{equation}
  \label{eqn:hadamard}
  (F[m]w)({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) w(t-\tau[m]({\bf x}_s,{\bf x}_r)) 
\end{equation}
In this approximation, and in other uses of the symbol $\approx$, the
error is $O(\lambda(w))$.

Approximation \ref{eqn:hadamard} follows from a
``progressing wave''  expansion of the fundamental
solution (Green's function) of the wave equation due to Hadamard
\cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
$m$, it holds for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
arrival'' condition fails (in
general) at larger distance. The precise region of validity depends on
$m$. For the remainder of this section, its validity is assumed.

Applying the approximation \ref{eqn:hadamard}, obtain
\begin{equation}
  \label{eqn:spwe}
S[m]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (w*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
\end{equation}
Assume that the data is noise-free:
$$
d({\bf x}_r,t;{\bf x}_s)=(F[m^*]w) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) w(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
$$
The (approximate) solution of $S[m]u[m;d]=d$ is obvious by inspection:
\begin{equation}
  \label{eqn:notl2}
  u[m;d]({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
\end{equation}
There is immediately a big problem: this $u[m;d]$ is not
square-integrable, and the AWI objective $J_{\rm AWI}$ is
undefined. To fix this, it's necessary to replace $u[m;d]$ by
$u_{0,\sigma}[m;d]$,  the solution of a regularized least squares problem: 
\begin{equation}
  \label{eqn:reg}
  u_{0,\sigma}[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
\end{equation}
$$
\approx \mbox{argmin}_u (\|a[m](w*u)(\cdot-\tau[m])-a[m^*]w(\cdot-\tau[m^*])\|^2 + \sigma^2\|u\|^2)
$$

Note that in these expressions, the source and receiver
coordinates ${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters
or summation indices. Therefore, to simplify the presentation, drop them from the notation
temporarily. Also,  use the temporary abbreviations $a^*=a[m^*], a=a[m], \tau^*=\tau[m^*],
\tau=\tau[m]$.

With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u_{0,\sigma}[m;d]$ and $w$:
\begin{equation}
  \label{eqn:freqreg}
\hat{u}[m;d] \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \sigma^2 |\hat{u}|^2\right)
\end{equation}
The normal equation for the least squares problem \ref{eqn:freqreg} is
\[
(a^2 |\hat{w}|^2 +\sigma^2)\hat{u} = a a^*|\hat{w}|^2e^{i\omega(\tau^*-\tau)}
\]
the solution of which is
\[
\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{\sigma}{a}} e^{i\omega(\tau^*-\tau)}
\]
where
$$ 
\hat{g}_{\sigma} = \frac{|\hat{w}|^2}{|\hat{w}|^2 + \sigma^2}.
$$
Transient acoustic sources are effectively bandlimited:
$\hat{w}(\omega) \rightarrow 0$ faster than any negative power of
$\omega$ as $|\omega| \rightarrow \infty$ and is entire analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$. Hence the inverse Fourier transform $g_{\sigma}$ tends to $\delta$ in the sense of distributions, and 
$$
u_{0,\sigma}[m;d](t) \approx \frac{a^*}{a}g_{\frac{\sigma}{a}}(t-(\tau^*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution
solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u_{0,\sigma}[m;d]$ is square integrable.

>>>>

While widely used in recent years, FWI exhibits a pathology when wave
velocities or their parametric equivalents are amongst the parameters
to be estimated: the excessive sensitivity of predicted waveforms to
changes in velocity leads to oscillations in the difference measures,
especially for the most commonly used square-integral measure. Local
optimization algorithms (relatives of Newton;'s method) are the only practical
methods to generate sequences of approximate optimal
parameters. However the oscillations in the difference measures lead
these sequences to become entrapped at near parameter distributions
far from the optimum, and thus do not extract the maximum information
from the data \cite[]{VirieuxOperto:09}.

@@

While these and other works discuss heuristic justifications for AWI and
many details of implementation, a clear explanation for its effectiveness is 
so far missing from the literature. In this short note, I show that
for active sensing of acoustic materials with unique arrival times of
transmitted waves
(that is, unique rays of geometric asymptotics between sources and
receivers), and known source wavelet, the AWI objective function is an asymptotic (in time
frequency) approximation to the
mean-square time-of-travel misfit, a commonly used objective for
travel time tomography. The error in this approximation is
proportional to the source mean-square wavelength. Thus AWI and travel time fitting
lead to similar material parameter estimates. Since the latter is
widely believed to be a near-optimal method for wave velocity
estimation, this relation explains roughly ``why AWI works'', at
least within this restricted scope.

@@

The version of AWI introduced by \cite{Warner:16} is based on an
active source acoustic model of data generation in Euclidean space
$\bR^3$. In this section, I will review a version of this model, place
some useful constraints on energy sources appearing in it, and define
AWI as an optimization problem.

\subsection{Acoustic Modeling and Waveform Inversion}

Each acoustic source is modeled an isotropic point radiator
with known location $\bx_s\in \bR^3$ and wave form (``wavelet'')
$w(t)$ (idealized as independent of $\bx_s$ for simplicity). These
sources generate acoustic waves traveling through Euclidean space. The
data are samples of these waves at receivers are modeled as isotropic
point sensors, recording (excess) pressure at points
${\bf x}_r \in \bR^3$. The recording takes place over a time interval
$[0,t_{\rm max}]$ (also idealized as independent of $\bx_s$ and
$\bx_r$) . I will refer to the recording interval and the
source-receiver position pairs collectively as the {\em acquisition
  geometry}.

The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$,
${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$
depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf
  x})$ (reciprocal of the density $\rho({\bf x})$), and source wavelet $w(t)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
                                    \left(\int^t w \right)
 \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The representation of the point source time-dependence as an
indefinite integral of the source wavelet will prove convenient
in expressing the asymptotic approximation of $p$.

For simplicity, I assume that the system \ref{eqn:awe} holds over all of Euclidean
space-time $(\bx,t) \in \bR^3 \times \bR$, and ignore physical
boundaries such as the Earth's surface.
For sake of brevity, define the model vector $m=(\kappa,\rho, w)$. The
{\em modeling operator} $F$ maps the model $m$ to the corresponding data $\{p({\bf x}_r,t;{\bf
  x}_s): 0 \le t \le t_{\rm max}\}$.

For the purposes of this paper, the source wavelet $w$ and the density
$\rho$ will be assumed known. In fact, even if actual acoustic sources are
well-approximated by isotropic point radiators, their wave forms are
usually poorly determined and may reasonably be
included as  parameters to be estimated by inversion of the
data. However for the present discussion, $w$ is regarded as
known.

Acoustic inversion means: given the set of
source-receiver position pairs $\{(\bx_s,\bx_r)\}$, source wavelet
$w(t)$, density $\rho(\bx)$,  and data traces
$\{d(\bx_r,t;\bx_s): 0 \le t \le t_{\rm max}\}$, find a bulk modulus $\kappa$ so that the model
$m=(\kappa,\rho,w)$ satisfies $F[m] \approx d$ in some sense. The
simplest version of FWI concretizes this task by asking for a model
$m$ (that is, a bulk modulus $\kappa$) minimizing the square integral/sum
error over the recording time interval $[0,t_{\rm max}]$ and
source-receiver position pairs:
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s}
  \int_0^{t_{\rm max}}\,dt\, |F[m]({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
\end{equation}
The definition \ref{eqn:fwi} makes sense only if the predicted and
observed data traces ($F[m]({\bf x}_r\cdot;{\bf x}_s)$ and $d({\bf
  x}_r,\cdot;{\bf x}_s)$) are square-integrable functions on $[0,t_{\rm max}]$, for each
source-receiver position pair $(\bx_r,\bx_s)$. This property is
assumed for $d$, and follows from additional assumptions on $m$ and the
acqquisition geometry, to be explained below.

It is convenient to introduce the $L^2$ norm symbol $\| \cdot \|$ as
shorthand for the right-hand side of the definition \ref{eqn:fwi}:
\[
  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]-d\|^2.
\]
In this notation, the interval $[0,t_{\rm max}]$ and the
source-receiver position pair set $\{(\bx_r,\bx_s)\}$ are implicit. In
the sequel, I will use the $L^2$ norm notation to stand for square
integral/sum over whatever variables make sense in context.

Practical versions of FWI include various types of regularization and
imposition of constraints on $m$, see the references cited in the
Introduction. As mentioned there, local iterative optimization methods
are the only practical approach to minimizing $J_{\rm FWI}$ due to the
computational cost of approximating the forward map $F$ and related
operators, but these tend to stagnate at highly sub-optimal
(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
quite high quality, often available only at considerable cost in data
collection and processing.

\subsection{Source Characteristics}

In applications to seismology, medical imaging, and other wave imaging
technologies, the wavelet $w$ is actually the product of signal
processing aimed to shape the data $F[m]$ into an approximation of
data that would have been predicted using a known wavelet. That is, the wavelet is to some extent {\em
  chosen}. This circumstance justifies several constraints on the
wavelet, consistent with the properties of observed signals: the
wavelet is assumed to be of class $H^1_0(\bR)$, that is, square integrable with
square inttegrable derivative and vanishing for sufficiently large
$t$, and even in $t$.
Introduce asymptotics in this setting by assuming that $w = \wl$ is chosen
from a family $\{\wl:\}$ indexed by the dimensionless parameter
$0 <\lambda <1\}$, and defined by
\begin{equation}
  \label{eqn:wfam}
  \wl(t) = \lambda^{-1/2}w_1(t/\lambda).
\end{equation}
The  ``mother wavelet'' $w_1$ is even and of class $H^1_0(\bR)$, and
all wavelets $\wl$ in this family
inherit these properties from $w_1$, along with its dimensions (energy/time$^2$).

It follows from these assumptions that the Fourier transform $\fwl$ is
real-valued, even, of class $H^1(\bR)$. and its product with frequency $|\omega|$ is also
square-integrable.

The parameter $\lambda$ is connected to several frequency-related
characteristics of $\wl$. Suppose that $f \in L^2(\bR)$ is also square-integrable after
multiplication by $t$ (true if  for instance $f$ has compact
support). The {\em pulse width} (or {\em RMS width} ) of $f$ is
\begin{equation}
  \label{eqn:pw}
  l(f) = \frac{\|tf\|}{\|f\|}.
\end{equation}
The pulse width of $\wl$ is proportional to $\lambda$:
\begin{equation}
  \label{eqn:wpw}
  l(\wl) = \left(\frac{\int dt (t \wl(t))^2}{\int dt (\wl(t))^2}
  \right)^{1/2} = \lambda l(w_1).
\end{equation}
The {\em RMS frequency} of $\wl$ is
\begin{equation}
  \label{eqn:freq}
  k(\wl) = \left(\frac{\int d\omega (\omega \fwl(\omega))^2}{\int dt (w(t))^2} \right)^{1/2} = \lambda^{-1}k(w_1).
\end{equation}
Thus the {\em RMS wavelength} $1/k(\wl)$ is proportional to $\lambda$.

The Heisenberg inequality (for example, \cite{Folland:07}, p. 255) bounds the RMS
wavelength, the reciprocal of the RMS frequency, in terms of the pulse
width:
\begin{equation}
  \label{eqn:heis}
  l(\wl) k(\wl) = l(w_1)k(w_1) \ge \frac{1}{4\pi}.
\end{equation}
Equality is attained for Gaussian $w_1$, and approximated for many other
commonly used wavelets (Ricker = second derivative of Gaussian,
trapezoidal bandpass,...).

\subsection{AWI via Optimization}

AWI is based an {\em extension} of the modeling operator $F$, that is,
an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
The extended modeling operator ${\bar F}$ maps extended models
$\bar{m}=(\kappa,\rho,\bar{w})$ incorporating extended sources
$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
field. That is, the extended source depends on the receiver location
as well as the source location - so there is one acoustic system for
each source {\em and} receiver position - a lot of wave equations! If all
of the wavelets for each source are the same, that is, ${\bar w}({\bf
  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
position, then ${\bar F}[\bar{m}] = F[m]$. That is, $F$ is a special
case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.

\cite{Warner:16} represent the extended sources $\bar{w}$ as time convolutions of the
(known) source $w$ with an {\em adaptive filter} $u({\bf
  x}_r,t;{\bf x}_s)$: that is,
\[
  \bar{w}({\bf x}_r,t;{\bf x}_s) = (u * w)({\bf  x}_r,t;{\bf x}_s)
  = (u({\bf  x}_r,\cdot;{\bf x}_s)*w)(t).
\]
Here, $u*w$ denotes the result of convolving each trace in $u$ by $w$. I
will also use the asterisk to denote the trace-by-trace convolution
of two data sets with the same set of source and receiver coordinates,
that is, the data set consisting of time convolutions of traces with the same source
and receiver coordinates. The meaning will be clear from context.

Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
\[
 \bar{F}[\bar{m}] = \bar{F}[(\kappa,\rho,,u*w)] = F[m]*u
\]
Introduce the notation $S[m]$ for the convolution operator with kernel $F[m]$:
\begin{equation}
  \label{eqn:sdef}
  F[m]*u \equiv S[m]u
\end{equation}
\cite{Warner:16} assume (implicitly) that there is always an exact
solution of the extended inversion problem. That is, they assume that
for any $m$, there is an extended source $\bar{w}$, or equivalently an
adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
$S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
around, as will be seen below. Also, \cite{Warner:16} formulate
AWI in terms of the inverse filter, that is, $u$ for which
$u*d = F[m]$, and in fact recommend that definition. The present
discussion uses only the forward filter.

If $\kappa = \kappa^*, \beta = \beta^*$, then the extended source
$\bar{w}$ that fits the data should be $= w$,
so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
${\bf x}_s,{\bf x}_r$. Such a kernel $u[m;d]$ is in the null space of
multiplication by $t$.
This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
\begin{equation}
  \label{eqn:tJz}
  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[m;d](\bx_r,t;\bx_s)|^2}{\int dt |u[m;d](\bx_r,t;\bx_s)|^2}\right)
\end{equation}
in which $u[m;d]$ is as defined above, that is, the solution of $S[m]u[m;d]
= d$.

Adaptive waveform inversion consists in iterative minimization of
$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta, w)$ of the
model, then update $m$ (that is, $\kappa$) repeatedly using a gradient
descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
and its gradient in detail, and their use in a quasi-Newton
optimization algorithm.

@@

It may not be obvious at first glance why the AWI approach to estimating
$m$, described in the last section, should be any more effective than
FWI. To see why AWI works when FWI does not, in at
least some settings, assume that the spatial dimension is three,
and make use of the geometric asymptotics (ray theory) approximation
to the solution of the wave equation. This approximation is {\em
  local}: it applies when sources and receivers are sufficiently
close. I will discuss its global failure and the consequences for AWI
in the fifth section of this paper.

Since the fields $\kappa$ and $\rho$ have been assumed smooth,
the geometric asymptotics approximation is applicable for solution of the point
radiator problem \ref{eqn:awe}. This approximation expresses the pressure field in terms
of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
travel time $\tau[m]({\bf x}_s,{\bf x}_r)$, and implies that
\begin{equation}
  \label{eqn:hadamard}
  F[m_{\lambda}]({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m]({\bf x}_s,{\bf x}_r)). 
\end{equation}
Here and in the notation to follow, $m_{\lambda} =
(\kappa,\rho,\wl)$.

In this approximation, and in other uses of the symbol $\approx$, the
error is dominated by data wavelength, that is, RMS wavelength of the
source wavelet $\wl$, which is proportional to $\lambda$ as mentioned above. In other words, ``$\approx$'' means ``in error by $O(\lambda)$''.

Approximation \ref{eqn:hadamard} follows from a
``progressing wave''  expansion of the fundamental
solution (Green's function) of the wave equation due to Hadamard
\cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
$m$, it holds {\em locally}, that is, for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
arrival'' condition fails (in
general) at larger distance. The precise region of validity depends on
$m$. For the remainder of this section, its validity is assumed.

Applying the approximation \ref{eqn:hadamard} to the definition
\ref{eqn:sdef}, obtain
\begin{equation}
  \label{eqn:spwe}
S[m_{\lambda}]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (\wl*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
\end{equation}
Assume that the data is noise-free:
$$
d({\bf x}_r,t;{\bf x}_s)=(F[m^*]) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
$$
The (approximate) solution of $S[m_{\lambda}]u=d$ is obvious by inspection:
\begin{equation}
  \label{eqn:notl2}
  u({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
\end{equation}
There is immediately a big problem: this $u$ is not
square-integrable, and the AWI objective $J_{\rm AWI}$ is
undefined. It's not surprising that mathematical ill-behaviour arises
in this context. Since $S[m_{\lambda}]$ is a convolution operator, solving
$S[m_{\lambda}]u=d$ for $u$ amounts to deconovolving $d$ by the kernel $F[m]$ of
$S[m_{\lambda}]$. Under the assumed conditions on the wavelet $\wl$, $S[m_{\lambda}]$ does
not have a continuous (stable) convolution inverse.

It is conventional to stabilize deconvolution problems such as the
equation $S[m_{\lambda}]u=d$ via so-called ``pre-whitening'', as
mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
Tihonov regularization. It  amounts to replacement of the equation
$S[m_{\lambda}]u=d$ by the regularized least squares problem: 
\begin{equation}
  \label{eqn:reg}
  u_0[m_{\lambda};d] = \mbox{argmin}_u (\|S[m_{\lambda}]u-d\|^2 + \lambda\|u\|^2)
\end{equation}

The choice of $\lambda$ as the  the regularization weight might seem
natural, as $\lambda$ is more-or-less the shortest wavelength
controlled by solution of the least-squares deconvolution problem
\ref{eqn:reg}. It will be rigorously justified {\em post-hoc} by the
calculations to follow.

Note that in these expressions, the source and receiver coordinates
${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters or
summation indices. The right-hand side of equation \ref{eqn:reg} is
summed over ${\bf x}_s$ and ${\bf x}_r$, for instance. However, the
quantities on the right-hand side are not coupled: the minimization
could be regarded as taking place independently for each choice of
${\bf x}_s$ and ${\bf x}_r$. Therefore, to simplify the presentation,
drop ${\bf x}_s$ and ${\bf x}_r$ from the notation temporarily: each
calculation to follow pertains to a single trace, i.e. a single choice
of ${\bf x}_s$ and ${\bf x}_r$.  Also, use the
temporary abbreviations
$a^*=a[m^*], a=a[m], \tau^*=\tau[m^*], \tau=\tau[m]$.

Using the Hadamard progressing wave approximation \ref{eqn:hadamard},
the right-hand side of equation \ref{eqn:reg} is 
$$
\approx \mbox{argmin}_u (\|a[m](\wl*u)(\cdot-\tau[m])-a[m^*]\wl(\cdot-\tau[m^*])\|^2 + \lambda\|u\|^2)
$$
Re-write the approximation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u_0}[m_{\lambda};d],\fwl$ of $u_0[m_{\lambda};d]$ and $\wl$:
\begin{equation}
  \label{eqn:freqreg}
\hat{u}_{0,\sigma}[m;d] \approx \mbox{argmin}_{\hat{u}}\frac{1}{2}\int d\omega \left(|a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \lambda |\hat{u}|^2\right)
\end{equation}
The normal equation for the least squares problem \ref{eqn:freqreg} is
\[
(a^2 |\fwl|^2 +\lambda)\hat{u} = a a^*|\fwl|^2e^{i\omega(\tau^*-\tau)}
\]
the solution of which is
\begin{equation}
  \label{eqn:hatuf}
\hat{u}_0[m_{\lambda};d] \approx \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
\end{equation}
where
\begin{equation}
  \label{eqn:hatgf}
\hat{g}_{\lambda,\sigma} = \frac{|\fwl|^2}{|\fwl|^2 + \sigma}.
\end{equation}
Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic,
hence vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$. Hence the inverse Fourier transform $g_{\lambda,\sigma}$ tends to $\delta$ in the sense of distributions, and 
$$
u_0[m_{\lambda};d](t) \approx \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution
solution of $\bar{S}[m_{\lambda}]u=d$ given in equation \ref{eqn:notl2}. However for $\lambda>0$, $u_0[m_{\lambda};d]$ is square integrable.
